---
title: HTH vs. HTT - The attempt of an explanation
author: Max Lang
date: '2021-04-25'
slug: hth-vs-htt-the-attempt-of-an-explanation
categories:
  - R
  - R Programming
  - Statistics
  - Mathematics
tags: []
---
# Question
This question was asked during the really interesting [TedTalk](https://www.ted.com/talks/peter_donnelly_how_juries_are_fooled_by_statistics?language=de) about statistics by Peter Donelly.

_Consider two Patterns HTH vs HTT. Which of the following is true?_

_a) The average number of tosses until HTH is larger than the average number of tosses until HTT_

_b) The average number of tosses until HTH is the same as the average number of tosses until HTT_

_c) The average number of tosses until HTH is smaller than the average number of tosses until HTT_

First one might think _b)_ is the correct answer. 
After a second thought one notices that the sequence _HTH_ overlaps itself. Let's say one got HTHTH it is clear that one got two occurrences of _HTH_ in only five coin tosses. (**HTH**TH and HT**HTH**).

Taking a look at the second sequence _HTT_ one will find out that there is not such an overlap.
Therefore one might think that _HTH_ is more likely than _HTT_ which is not the case as shown in the code chunk below.

1. One should note that we are either looking for _HTT_ or (exclusive) for _HTH_. 

2. We only flip until the pattern which we are looking for got hit.

```{r}
set.seed(10)

fair_coin <- c('H','T')

seek <- function(seq) {
  # Flipping the coin three times
    result <- sample(x= fair_coin, size= 3,replace=TRUE)
  # Defining the minimum Number of flips (three)
    flips <- 3
  # Setting  issed to true so while loop iterates at least one time
    missed <- TRUE
  
    while (missed == TRUE) {
      # If all our elements in the vector are the same as the sequence we seek stop the loop
        if (all(seq == result)) {
            missed <- FALSE
        }
      # Recycle last two elements and flip again once (important)
        else {
            result <- c(result[2],result[3], sample(x= fair_coin, size= 1, replace= TRUE))
            flips <- flips + 1
        }
    }
    return(flips)
}
# Setting the number of flips of t experi
n <- 10000

experiment <- data.frame("HTH"=rep(NA,n),"HTT"=rep(NA,n))

HTH <- c('H','T','H')
HTT <- c('H','T','T')

for (i in 1:n) {
    experiment[i,] <- c(seek(HTH),seek(HTT))    
}
# Average number of flips
summary(experiment)
```

## The attempt of an explanation:

Let's suppose the following scenarios.

#### Looking for the sequence _HTH_

*HTH*

The simplest scenario in which we win right away.

*HTT*

Because of the last _T_ in the sequence one needs to start tossing until one gets another _H_ to "start" the _HTT_ sequence.

_Comment 1_: 

Let's suppose one already got _HT_. If the next toss is _H_ one found the sequence and is done. If it's a _T_,  one has to start all over again: since the last two tosses were _TT_ one now needs the full _HTH_.

#### Looking for _HTT_

*HTT*

Again the simplest scenario in which we win right away.

*HTH*

Because we are looking for _HTT_ we did not win in this case. However, and this is the important, at this stage one does not need to toss the coin again in order to find the starting token _H_.

_Comment 2_: 

Let's suppose again one already got _HT_. If the next toss is _T_, one found the sequence. If it is a _H_, this is just a minor setback, because one now got _HTH_ (while seeking the _HTT_ sequence); however, now one already has the _H_ and just needs two _T_ in a row ( _TT_ ). Moreover if the next toss is an H this is not such a big setback again, because _H_ is the start of the sequence we are looking for, just like before. A _T_ would make our "situation" even better and so on... 

Therefore answer _a) The average number of tosses until HTH is larger than the average number of tosses until HTT_.

## Experiment 2

After we put in some thought, we wanted to find how the probabilities of those two sequences behave if we adjust the experiment.

Now we will throw a coin 100 Times and note down each outcome. The result of each run of the experiment will therefore be a line/string looking e.g like this  one "HTHTTTHHHTHTHTHTHT...", containing in total 100 characters. After this we go through our 100 character long string and check for the following patterns. 

* **HTH**
* **HTT**

Note that HTH can overlap itself. For example _HTHTH_ is counted as **HTH**TH and HT**HTH** -  two sequences for *HTH*. **HTT** does not have this characteristic. 

The **Question** is which of the following statements is true:

* The number of times **HTH** is in our string is more than the number of times **HTT** occurs
* The number of times **HTH** is in our string is less than the number of times **HTT** occurs
* The number of times **HTH** is in our string is equal to the number of times **HTT** occurs

Let's make a simple simulaton in R to find out the correct answer. We first defined this function to simulate or experiment.
```{r}
count.pattern <- function(pattern,
                          n_flips = 100,
                          coins = rbinom(n_flips, 1, 0.5)) {
  hits = 0
  for (i in 1:(n_flips - 2)) {
    # TRUES are counted as 1 in R, this allows the following step
    if (sum(coins[i:(i + 2)] == pattern) == 3) {
      hits = 1 + hits
    }
  }
  hits
}
```


### Run the experiment 
```{r}

# Encoded 0 = H and 1 = T
HTH = c(0, 1, 0)
HTT = c(0, 1, 1)
TTT = c(1, 1, 1)

n_flips =  100

# Running the experiment 10000 times seeking HTH, HTT and TTT
results = data.frame(
  HTH = replicate(10000, count.pattern(HTH, n_flips)),
  HTT = replicate(10000, count.pattern(HTT, n_flips)),
  TTT = replicate(10000, count.pattern(TTT, n_flips))
)
summary(results)[4,]

```
The **means** at 10000 trials are close enough that we can safely assume that
the average amount of successes **is the same**, no matter which pattern.

However, their **distribution is not**:

### Plotting the histograms of all three patterns
```{r}
ggplot(tidyr::pivot_longer(results,
                           c("HTH", "HTT", "TTT"),
                           names_to = "pattern",
                           values_to = "value"),
       aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(vars(pattern)) +
  labs(x = paste("Amount of pattern occurences.")) +
  ggtitle(paste("Distribution of different pattern occurrences in", n_flips, "coin tosses."))

```


### Also interesting: 5 tosses

The means stay the same, but at this few tosses, some patterns can only appear
once, while othes (TTT) can appear up to three times when overlapping is allowed.
```{r}
n_flips =  5

# Running the experiment 10000 times seeking HTH, HTT and TTT
results = data.frame(
  HTH = replicate(50000, count.pattern(HTH, n_flips)),
  HTT = replicate(50000, count.pattern(HTT, n_flips)),
  TTT = replicate(50000, count.pattern(TTT, n_flips))
)
print(summary(results)[4,])

ggplot(tidyr::pivot_longer(results,
                           c("HTH", "HTT", "TTT"),
                           names_to = "pattern",
                           values_to = "value"),
       aes(value)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(vars(pattern)) +
  labs(x = paste("Amount of pattern occurences.")) +
  ggtitle(paste("Distribution of different pattern occurrences in", n_flips, "coin tosses."))
```